---
title: "Processing ARC AIRR"
author: "Peter Fiorica"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

As part of the ARC Project, I was asked to calculate adaptive immune receptor repertoire signatures. The plan is to use MiXCR as I did for the TILS project to calculate adaptive immune receptor repertoire information. The details of how I did this in the past can be found at `/projects/rpci/songyao/pnfioric/tils_r01_RNAseq/tcr_quant/00_MiXCR_Introduction.Rmd`.

Similar to last time, we will be using the software [badger](https://github.com/hakyimlab/badger) to call submit the MiXCR job in an HPC environment since we have 1000+ samples. The samples are store in the directories below. The plan will be to harmonize these paths with symbolic links to be able to call all of these jobs with badger.

```
#RQ-026109 plates 1-4
/projects/rpci/songliu/liyan/LiYan/RQ-026109_1to4/RQ026109-Yao/
#192 Samples

#RQ-026109 plates 5-8
#/projects/rpci/zhihongg/CapRun/RQ026109_Yao/ # This was replaced by the directory below...
/projects/rpci/songliu/liyan/LiYan/RQ-026109_5to8/RQ026109_Yao/
# 192 Samples
#RQ-026110 plates 9-12
/projects/rpci/songliu/liyan/LiYan/RQ-026110-Yao/
# 192 Samples
#RQ-026110 plates 13-16
/projects/rpci/songliu/liyan/LiYan/RQ-026110_13to16/RQ026110-Yao/
#192 Samples
#RQ-026111 plates 17-18
/projects/rpci/songliu/liyan/LiYan/RQ026111/RQ026111-Yao/
#96 Samples
```

# Harmonizing File Paths

```
bash harmonize_symbolic_links.sh
```

# Badger Files

Badger uses `.yaml` and `.jinja` job templates to submit jobs. In our situation, we will have to edit the previous files to submit these jobs, but hopefully only the `.yaml` file will need to be edited. We are commenting out the `#fake_submission: true` piece, so the jobs will not be submitted immediately after they are generated. Since we are generating >1000 jobs, this would stall the CCR queue, so instead will be using a simple script to submit the jobs at regularly spaced intervals.

Files with the FASTQ files are named as such:

```
RS-04367378_VALS11004032-A5_RS-03837571_S44_R1_001.fastq.gz
RS-04367378_VALS11004032-A5_RS-03837571_S44_R2_001.fastq.gz
```

In these, the first RS_ID (RS-04367378) corresponds to the ARC coldata, so we will be abstracting that from the file name.. Our code is set up to run this way in the badger `call_mixcr.yaml`

We can run this with 

```
python /projects/rpci/songyao/pnfioric/software/badger/src/badger.py -yaml_configuration_file /projects/rpci/songyao/pnfioric/arc_project/airr_calculation/call_mixcr.yaml -parsimony 8
```

It looks like this code prepared all of the jobs appropriately above. We input 864 files and have 864 jobs. 

*Note:* I noticed that in the previous ARC data from DE-analysis, there appears to be 1112 samples in the working dataset--Of these samples, 687 are from ARC, so I am wondering where this 864 comes from.


#### Initial Error

When I was initially running MiXCR with these data, I received the error:

```
java.lang.RuntimeException: java.io.IOException: No locks available
        at com.milaboratory.o.cG.a(SourceFile:137)
        at com.milaboratory.o.cJ$a.run(SourceFile:484)
Caused by: java.io.IOException: No locks available
        at java.base/sun.nio.ch.FileDispatcherImpl.lock0(Native Method)
        at java.base/sun.nio.ch.FileDispatcherImpl.lock(FileDispatcherImpl.java:96)
        at java.base/sun.nio.ch.FileChannelImpl.lock(FileChannelImpl.java:1119)
        at java.base/java.nio.channels.FileChannel.lock(FileChannel.java:1063)
        at com.milaboratory.o.cG.a(SourceFile:116)
        ... 1 more
License manager thread died.
License error: UnknownError
```

LLMs did not have a good response to this that was feasible, so I explored the MiXCR github since I had not received the error in the past. The discussion is documented at: https://github.com/milaboratory/mixcr/issues/1923 .


I had to [request a new academic license](https://platforma.bio/getlicense/) from mixcr since my previous one had expired.

I successfully activated the new license and it appears it runs fine now.

# Immunarch

### Diversity Processing

```{r, echo = FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(readxl))
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(gridExtra))
library(immunarch)
"%&%"=function(a,b) paste(a,b,sep="")

```

```{r load_demographics}
arc_dir <- "/projects/rpci/songyao/pnfioric/arc_project/"

base_path <- "/projects/rpci/songyao/pnfioric/arc_project/airr_calculation/tcr_quant/results"

file1<-read_xlsx(arc_dir %&% "ER Backfill AsA Only RNAseq clinical data.xlsx") %>% rename(Sample=FASTQ_RS)
fwrite(file1, base_path %&% "metadata.txt", col.names = T, sep = "\t", row.names = F, quote = F)

#file2 <- read_xlsx(arc_dir %&% "Updated ARC RNAseq clinical data 02-25-2026.xlsx")
```

The file we are reading above here as `file1` is a broader sample information file that includes information to the demographics and processing of these samples. This file includes `r nrow(file1)` samples in it. All of these samples are from the ARC study, and it appears there are no validation samples in this group.

*Notably*, in there are 865 FASTQ samples that I processed with MiXCR, so not all of them will overlap.

```{r}
protein_products <- c("IGH", "IGK", "IGL", "TRA", "TRB", "TRD", "TRG")

immdata_list <- list()
for (chain in protein_products) {

  files <- list.files(
    path = base_path,
    pattern = paste0("clones_", chain, "\\.tsv$"),
    full.names = TRUE
  )
  cat("Found", length(files), "files for", chain, "\n")
  if (length(files) == 0) next
  immdata <- tryCatch(
    repLoad(files),
    error = function(e) {
      message("Failed loading ", chain)
      message(e)
      NULL
    }
  )
  if (!is.null(immdata)) {
    object_name <- paste0("immdata_", chain)
    assign(object_name, immdata)
    immdata_list[[object_name]] <- immdata
  }
}

```

# Loop over Diversity Calculations

```{r, warning= FALSE}
suffixes <- protein_products

# Initialize empty lists to store results
div_hill_list <- list()
div_d50_list <- list()
div_div_list <- list()
div_gini_simp_list <- list()
div_raref_list <- list()

# Loop over each suffix
for (suf in suffixes) {
  # Construct the variable name dynamically
  immdata_name <- paste0("immdata_", suf)
  
  # Check if the object exists
  if (exists(immdata_name)) {
    immdata_obj <- get(immdata_name)  # Retrieve the object

    # Calculate diversity measures
    div_hill_list[[suf]] <- repDiversity(immdata_obj$data, "hill")
    div_d50_list[[suf]] <- repDiversity(immdata_obj$data, "d50")
    div_div_list[[suf]] <- repDiversity(immdata_obj$data, "div")
    div_gini_simp_list[[suf]] <- repDiversity(immdata_obj$data, "gini.simp")
    #div_raref_list[[suf]] <- repDiversity(immdata_obj$data, "raref")
  } else {
    warning(paste("Object", immdata_name, "not found. Skipping."))
  }
}
# The results are stored in:
# div_hill_list: Hill numbers for each suffix
# div_d50_list: D50 diversity for each suffix
# div_div_list: Ecological diversity for each suffix

# Example: Accessing the Hill numbers for IGH
#div_hill_list[["IGH"]]

```

## Convert all to Data.tables

```{r}
div_hill_df <- bind_rows(lapply(names(div_hill_list), function(suf) {
  df <- as.data.frame(div_hill_list[[suf]])  # Ensure it's a data frame
  df$suffix <- suf  # Add suffix column
  return(df)
}), .id = "dataset")

div_d50_df <- bind_rows(lapply(names(div_d50_list), function(suf) {
  df <- as.data.frame(div_d50_list[[suf]])  # Ensure it's a data frame
  df$suffix <- suf  # Add suffix column
  return(df)
}), .id = "dataset")
div_d50_df$Sample <- rownames(div_d50_df)
div_d50_df<- div_d50_df %>% rename(Value=Clones)

div_div_df <- bind_rows(lapply(names(div_div_list), function(suf) {
  df <- as.data.frame(div_div_list[[suf]])  # Ensure it's a data frame
  df$suffix <- suf  # Add suffix column
  return(df)
}), .id = "dataset")
#div_div_df$Sample <- rownames(div_div_df)

div_gini_simp_df <- bind_rows(lapply(names(div_gini_simp_list), function(suf) {
  df <- as.data.frame(div_gini_simp_list[[suf]])  # Ensure it's a data frame
  df$suffix <- suf  # Add suffix column
  return(df)
}), .id = "dataset")
#div_div_df$Sample <- rownames(div_div_df)

#div_raref_df <- bind_rows(lapply(names(div_raref_list), function(suf) {
#  df <- as.data.frame(div_raref_list[[suf]])  # Ensure it's a data frame
#  df$suffix <- suf  # Add suffix column
 # return(df)
#}), .id = "dataset")

all_div_df <- bind_rows(
  #div_hill_df %>% mutate(Diversity = "Div_Hill"),
  div_d50_df %>% mutate(Diversity = "Div_D50"),
  div_div_df %>% mutate(Diversity = "Div_Div"),
  div_gini_simp_df %>% mutate(Diversity = "Div_GiniSimp")
)

all_div_wide <- all_div_df  %>% 
  mutate(Sample=gsub("mixcr_", "", Sample)) %>% 
  mutate(Sample=gsub(".clones.*","", Sample)) %>%
  pivot_wider(
    id_cols = Sample,
    names_from = c(Diversity, suffix),
    names_sep = "_",
    values_from = Value
  )

fwrite(all_div_wide, arc_dir %&% "airr_calculation/results/all_div_calculations.csv", row.names = F, col.names = T, sep = ",", quote = F )
fwrite(all_div_wide, arc_dir %&% "airr_calculation/results/all_div_calculations.txt", row.names = F, col.names = T, sep = "\t", quote = F )
```

## Hill Diversity

```{r}
div_hill_df$Sample <- gsub(".clones.*","", div_hill_df$Sample)
div_hill_df_meta <- left_join(div_hill_df, file1, by = "Sample") %>% filter(!is.na(ER))

igh_meta <- immdata_IGH$meta
igh_meta$RS_ID <- str_extract(igh_meta$Sample, "RS-\\d+")
igh_meta <- left_join(igh_meta, file1, by = c("RS_ID"= "Sample"))
igh_meta$Sample <-  gsub(".clones.*","", igh_meta$Sample)

div_hill_df$Sample <- gsub(".clones.*","", div_hill_df$Sample)
div_hill_df_meta <- left_join(div_hill_df, igh_meta, by = "Sample") %>% filter(!is.na(ER))

fwrite(div_hill_df_meta, arc_dir %&% "airr_calculation/results/div_hill_data_mixcr.txt", col.names = T, row.names = F, sep = "\t", quote = F)


ggplot(div_hill_df_meta, aes(x = as.factor(Q),
                             y = Value,
                             fill = interaction(ER, race))) +
  geom_boxplot() +
  theme_bw() + facet_wrap(~suffix, scales="free_y", nrow=1)
```

## Div D50 List

```{r}
div_d50_df$Sample <- gsub(".clones.*","", div_d50_df$Sample)
div_d50_df_meta <- left_join(div_d50_df, igh_meta, by = "Sample") %>% filter(!is.na(ER))

fwrite(div_d50_df_meta, arc_dir %&% "airr_calculation/results/div_d50_data_mixcr.txt", col.names = T, row.names = F, sep = "\t", quote = F)

ggplot(div_d50_df_meta, aes(x = interaction(ER, race),
                             y = Value,
                             fill = interaction(ER, race))) +
  geom_boxplot() +
  theme_bw() + facet_wrap(~suffix, scales="free_y", nrow=1)
```


## Ecological Diversity

```{r}
div_div_df$Sample <- gsub(".clones.*","", div_div_df$Sample)
div_div_df_meta <- left_join(div_div_df, igh_meta, by = "Sample") %>% filter(!is.na(ER))

fwrite(div_div_df_meta, arc_dir %&% "airr_calculation/results/div_diversity_eco_data_mixcr.txt", col.names = T, row.names = F, sep = "\t", quote = F)

ggplot(div_div_df_meta, aes(x = interaction(ER, race),
                             y = Value,
                             fill = interaction(ER, race))) +
  geom_boxplot() +
  theme_bw() + facet_wrap(~suffix, scales="free_y", nrow=1) #+ stat_compare_means(method = "anova", label = "p.format") 
```

# Gini-Simpson Index

```{r}
div_gini_simp_df$Sample <- gsub(".clones.*","", div_gini_simp_df$Sample)
div_gini_simp_df_meta <- left_join(div_gini_simp_df, igh_meta, by = "Sample") %>% filter(!is.na(ER))

fwrite(div_gini_simp_df_meta, arc_dir %&% "airr_calculation/results/div_gini_simp_data_mixcr.txt", col.names = T, row.names = F, sep = "\t", quote = F)
```

# Combine into one file

```{r}
#
div_values <- div_div_df_meta %>%
  select(Sample, Div_Div=Value, suffix, RS_ID) %>%
  pivot_wider(
    id_cols = c(Sample, RS_ID),
    names_from = suffix,
    values_from = Div_Div,
    names_prefix = "Div_Div_"
  )

d50_values <- div_d50_df_meta %>%
  select(Sample, Div_D50=Clones, suffix, RS_ID )%>%
  pivot_wider(
    id_cols = c(Sample, RS_ID),
    names_from = suffix,
    values_from = Div_D50,
    names_prefix = "Div_D50_"
  )

gini_values <- div_gini_simp_df_meta %>%
  select(Sample,Div_GiniSimp=Value, suffix, RS_ID) %>%
  pivot_wider(
    id_cols = c(Sample, RS_ID),
    names_from = suffix,
    values_from = Div_GiniSimp,
    names_prefix = "Div_GiniSimp_"
  )

```


```{r}
library(dplyr)
library(parallel)

# Create a named list of your immdata_* objects
immdata_list <- list(
  IGH = immdata_IGH,
  IGK = immdata_IGK,
  IGL = immdata_IGL,
  TRA = immdata_TRA,
  TRB = immdata_TRB,
  TRD = immdata_TRD,
  TRG = immdata_TRG
)

# General processing function
process_chain <- function(chain_name, immdata_obj) {
  df <- immdata_obj$meta
  
  # Extract base ID
  df$BaseID <- gsub(paste0("\\.clonotypes\\.", chain_name), "", df[,1])
  
  # Identify clonotype rows for this chain
  is_clonotype_row <- grepl(paste0("clonotypes\\.", chain_name), df[,1])
  
  sample_rows <- df[!is_clonotype_row, ]
  clonotype_rows <- df[is_clonotype_row, ]
  
  clonotype_rows <- clonotype_rows[, c("BaseID"), drop=FALSE]
  
  df_clean <- merge(sample_rows, clonotype_rows, by="BaseID", all.x=TRUE)
  
  df_clean$ClonotypeID <- paste0(df_clean$BaseID, ".clonotypes.", chain_name)
  df_clean$Sample <- df_clean$ClonotypeID
  
  df_clean <- df_clean %>% select(-BaseID, -ClonotypeID)
  
  # Overwrite the meta data inside the immdata object
  immdata_obj$meta <- df_clean
  
  return(immdata_obj)
}

# Now apply across all chains
processed_immdata_list <- mapply(process_chain, names(immdata_list), immdata_list, SIMPLIFY = FALSE)

# You now have:
# processed_immdata_list$IGH
# processed_immdata_list$TRB
# etc.

```

# Basic Clonality

## Volume

```{r, volume}
#group_levels <- levels(interaction(meta$race, meta$ER, drop = TRUE))
#colors_pal <- colorRampPalette(c("#68c766", "#a9e3a7",  "#9c6ba6", "#420353"))(length(group_levels))

plot_list_race_er <- list()

# Loop through each chain
for (chain in protein_products) {
  #message("Processing: ", chain)

  data <- processed_immdata_list[[chain]]$data
  meta <- processed_immdata_list[[chain]]$meta
  meta <- meta %>% filter(!is.na(race), !is.na(ER))

  # Run repExplore
  exp_vol <- repExplore(data, .method = "volume", .coding = T)

  # Visualization with two groupings
  p2 <- vis(exp_vol, .by = c("race", "ER"), .meta = meta) + ggtitle(paste(chain, "- Race x ER"))

  plot_list_race_er[[chain]] <- p2
  print(p2)
}
```

## Length

```{r, fig.width =18}
plot_list_race_er <- list()

for (chain in protein_products) {
  data <- processed_immdata_list[[chain]]$data
  meta <- processed_immdata_list[[chain]]$meta
  
  # Skip if data is NULL or not a list/data.frame
  if (is.null(data) || length(data) == 0 || all(sapply(data, is.null))) {
    message(paste("Skipping", chain, ": no usable data"))
    next
  }

  # Filter meta
  meta <- meta %>% filter(!is.na(race), !is.na(ER))
  
  # Optionally remove samples with no clonotypes
  data <- data[sapply(data, function(x) !is.null(x) && nrow(x) > 0)]
  
  if (length(data) == 0) {
    message(paste("Skipping", chain, ": no clonotypes in any sample"))
    next
  }

  # Run repExplore and plot
  exp_vol <- repExplore(data, .method = "len", .col = "aa")
  p2 <- vis(exp_vol, .by = c("race", "ER"), .meta = meta, .width=0) +
    ggtitle(paste(chain, "- Race x ER"))
  
  plot_list_race_er[[chain]] <- p2
  print(p2)
}


```

## Count

```{r}
plot_list_race_er <- list()

for (chain in protein_products) {
  data <- processed_immdata_list[[chain]]$data
  meta <- processed_immdata_list[[chain]]$meta
  
  # Skip if data is NULL or not a list/data.frame
  if (is.null(data) || length(data) == 0 || all(sapply(data, is.null))) {
    message(paste("Skipping", chain, ": no usable data"))
    next
  }

  # Filter meta
  meta <- meta %>% filter(!is.na(race), !is.na(ER))
  
  # Optionally remove samples with no clonotypes
  data <- data[sapply(data, function(x) !is.null(x) && nrow(x) > 0)]
  
  if (length(data) == 0) {
    message(paste("Skipping", chain, ": no clonotypes in any sample"))
    next
  }

  # Run repExplore and plot
  exp_vol <- repExplore(data, .method = "count")
  p2 <- vis(exp_vol, .by = c("race", "ER"), .meta = meta) +
    ggtitle(paste(chain, "- Race x ER"))
  
  plot_list_race_er[[chain]] <- p2
  print(p2)
}
```

# Clonality

```{r, eval = FALSE}
imm_pr_total <- data.table()
imm_hom_total <- data.table()

for (chain in protein_products) {
  data <- processed_immdata_list[[chain]]$data
  meta <- processed_immdata_list[[chain]]$meta
  
  # Skip if data is NULL or not a list/data.frame
  if (is.null(data) || length(data) == 0 || all(sapply(data, is.null))) {
    message(paste("Skipping", chain, ": no usable data"))
    next
  }

  # Filter meta
  meta <- meta %>% filter(!is.na(race), !is.na(ER))
  
  # Optionally remove samples with no clonotypes
  data <- data[sapply(data, function(x) !is.null(x) && nrow(x) > 0)]
  
  if (length(data) == 0) {
    message(paste("Skipping", chain, ": no clonotypes in any sample"))
    next
  }

  # Run repExplore and plot
  imm_pr <-  repClonality(data, .method = "top", .head = c(10, 100, 1000, 3000, 10000))
  imm_pr$suffix <- chain
  
  
  imm_hom <- repClonality(data,
      .method = "homeo",
      .clone.types = c(Small = .0001, Medium = .001, Large = .01, Hyperexpanded = 1)
    )
  imm_hom$suffix <- chain
  
  imm_pr_total <- bind_rows(imm_pr_total, imm_pr)
  imm_hom_total <- bind_rows(imm_hom_total, imm_hom)

}
```

# Gene usage

```{r}
# Loop over each chain in your processed_immdata_list
for (chain in names(processed_immdata_list)) {
  cat("Processing:", chain, "\n")
  # Run pubRep on each chain, using "aa+v" and coding sequences only
  rep_list[[chain]] <- pubRep(
    processed_immdata_list[[chain]]$data,
    "aa+v",
    .verbose = TRUE,
    .coding = TRUE
  )
}
```


```{r, eval = FALSE}
pr.trb <- pubRep(immdata_TRB$data, "aa+v", .verbose = T, .coding = T)
db <- dbLoad(file_path, "vdjdb", "HomoSapiens", "TRB")

```

```{r}
pubrep_list <- list()

# Loop over each chain in your processed_immdata_list
for (chain in names(processed_immdata_list)) {
  cat("Processing:", chain, "\n")
  # Run pubRep on each chain, using "aa+v" and coding sequences only
  pubrep_list[[chain]] <- pubRep(
    processed_immdata_list[[chain]]$data,
    "aa",
    .verbose = TRUE,
    .coding = TRUE
  )
}

```

```{r}
pubrep_list_v <- list()

# Loop over each chain in your processed_immdata_list
for (chain in names(processed_immdata_list)) {
  cat("Processing:", chain, "\n")
  # Run pubRep on each chain, using "aa+v" and coding sequences only
  pubrep_list_v[[chain]] <- pubRep(
    processed_immdata_list[[chain]]$data,
    "aa+v",
    .verbose = TRUE,
    .coding = TRUE
  )
}
```


```{r}
vdjdb = dbLoad("https://gitlab.com/immunomind/immunarch/raw/dev-0.5.0/private/vdjdb.slim.txt.gz", "vdjdb", .species = "HomoSapiens")
```

```{r}
pr1 <- pubRepFilter(pubrep_list$TRA, processed_immdata_list$TRA$meta, c(race = "Black"))
pr2 <- pubRepFilter(pubrep_list$TRA, processed_immdata_list$TRA$meta, c(race = "White"))


pr3 <- pubRepApply(pr1, pr2)
```


```{r}
pr1 <- pubRepFilter(pubrep_list$IGH, processed_immdata_list$IGH$meta, c(race = "Black"))
pr2 <- pubRepFilter(pubrep_list$IGH, processed_immdata_list$IGH$meta, c(race = "White"))


pr3 <- pubRepApply(pr1, pr2)
```

